How to improve your relationship with your future self

### 2016-12-13 workflow.Rmd
Edited up to Meaningful Code..

---
title: How to improve your relationship with your future self
author:
- name: Jake Bowers
  affiliation: <jwbowers@illinois.edu>,  University of Illinois at Urbana-Champaign, Departments of Political Science and Statistics; Fellow, White House Social and Behavioral Sciences Team.
- name: Maarten Voors
  affiliation: <maarten.voors@wur.nl>, Wageningen University, Development Economics Group.  \emph{Acknowledgements:} Many thanks to the EGAP Learning Days 2016 participants in Santiago Chile, to the BITSS team, the Department of Economics and Ted Miguel at UC Berkeley where Maarten was a visiting researcher during spring 2016. Maarten gratefully acknowledges financial support from N.W.O. grant 451-14-001. A previous version of this paper benefited from comments and discussions with Mark Fredrickson, Brian Gaines, Kieran Healy, Kevin Quinn, Cara Wong, Mika LaVaque-Manty and Ben Hansen. The source code for this document may be freely downloaded and modified from <https://github.com/jwbowers/workflow>. This paper extends a previous version by Jake @bowers2011tpm. While some text remains the same we have expanded and updated the essay to reflect current developments.
bibliography: references.bib
csl: american-political-science-review.csl
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
    pdf_document:
      keep_tex: true
      number_sections: true
      fig_width: 7
      fig_height: 7
      fig_caption: true
      template: bowersarticletemplate.latex
      pandoc_args: [
      "-M", "secPrefix=\\S",
      "--filter", "/usr/local/bin/pandoc-crossref"
      ]
      md_extensions: +autolink_bare_uris+ascii_identifiers+tex_math_single_backslash
...

<!-- An example comment -->
<!-- Separate the thanks sections from affiliation and integrate? Wasnt sure how to do this here-->

@@

## To make the html file do
## render("exploration4.Rmd",output_format=html_document(fig_retina=FALSE))
## To make the pdf file do
## render("exploration4.Rmd",output_format=pdf_document())
## to debug, try render("thefile.Rmd", clean=FALSE) so that you can see the intermediate files

require(knitr)
opts_chunk$set(
  tidy=TRUE,     # display code as typed
  size="footnotesize",    # slightly smaller font for code
  echo=TRUE,
  results='markup',
  strip.white=TRUE,
  fig.path='figs/fig',
  cache=FALSE,
  highlight=TRUE,
  width.cutoff=100,
  out.width='.9\\textwidth',
  message=FALSE,
  comment=NA)
```

```{r initialize,echo=FALSE}
##First, just setup the R environment for today:
if(!file.exists('figs')) dir.create('figs')

options(SweaveHooks=list(fig=function(){
			   par(mar=c(3.5, 3, 1.1, 0),
			       pty="s",
			       mgp=c(1.5,0.5,0),
			       oma=c(0,0,0,0))},
			 echo=function(){options(continue=" ") ##Don't show "+" prompts,
			 options(prompt=" ")
			 }),
	digits=4,
	scipen=8,
	width=100
	)

#options(error=function(){options(prompt="> ",continue="+ ");NULL})
```

@@

Memory is tricky. Learning requires effort. When we do not practice and repeat something that we want to remember, most people forget quickly, are overconfident in their abilities to recall future information [@koriat2005illusions], and may even recall events that never happened.^[See the following site for a nice overview of what we know about memory including the fact that learning requires practice: <http://www.spring.org.uk/2012/10/how-memory-works-10-things-most-people-get-wrong.php>. On false memory see Wikipedia and linked studies <https://en.wikipedia.org/wiki/False_memory>.] Moreover, most of us live busy lives. We write text and code behind our new standing desk at home while the kids run around, cry about an elder sister taking away a favorite new yellow school bus toy, drinks spill on the floor, the laundry needs doing, the kettle boils \ldots Our lives and minds are full and we need to efficiently move from task to task. If we cannot count on memory, then how can we do science?

How long does it take from planning a study to publication? How much time passes to the first reproduction of a study? Is three years too short? Is ten years too long? We suspect that few of our colleagues in the social and behavioral sciences conceive of, field, write, and publish a data driven study faster than about three years. We also suspect that, if some other scholar decides to re-examine the analyses of a published study, it will occur after publication. Moreover, this new scholarly activity of learning from one anothers' data and analyses, can occur at anytime, many years past the initial publication of the article.^[The process of reproducing past findings can occur when one researcher wants to build on the work of another. It can also occur within the context of classes --- some professors assign reproduction tasks to students to aid learning about data analysis and statistics. In addition to those models, recently reproduction of research has been organized to enhance the quality of public policy in the field of economic development by the 3IE Replication Program (@brown2014quality) and to assess the quality of scientific research at the level of sub-disciplines in social psychology (@open2015estimating) and within experimental economics (@camerer2016evaluating). In another study 29 research teams recently collaborated on a project focusing on applied statistics to see if the same answers would emerge from re-analyses of the same data set (@silberzahn2015crowdsourced).]

@@

```{r read.data,echo=FALSE,results='hide', cache=TRUE}
## This data frame will be used throughout the analyses.
## Putting it up top, so we can use it in the comments section.
library(readstata13)
dem.nations.df<-read.dta13("https://www.dropbox.com/s/huhqag5rudwtbno/Democracy%20Cross-National%20Data%20V4.1%2009092015.dta?dl=1&pv=1&raw=1",convert.factors=FALSE)

row.names(dem.nations.df)<-dem.nations.df$Nation ##make the row names more helpful
good.df<-na.omit(dem.nations.df[,c("Gini2004","WVS5_protest","Nation","meanpr")])
good.df$gini04<-good.df$Gini2004/100
good.df$protest05<-good.df$WVS5_protest
```

We organize the paper around a series of homilies that lead to certain concrete actions.

@@

All your results (numbers, comparisons, tables, plots, figures) should be produced from
code, not from a series of mouse clicks or copying and pasting.^[To our future
selves: both of us are using a computer control device called a track-pad, and
haven't used an older device called a
[mouse](https://en.wikipedia.org/wiki/Computer_mouse) in some years. We no
longer hear a clicking sound when we operate the control device either. We are
not sure why we talked about mouse clicks just now.] Imagine you wanted to re-create
a figure and include a new variable, you should be able to do so with just a few edits to the code rather than knowledge
of how you used a pointing device in your graphical user interface all those
years ago.

Lets look at an example. Using R [@R:2011], a open-source statistical software program, you
might specify that a file, called `fig1.pdf` is produced by the following set
of commands in a file called `makefig1.R`. Lets look at some annotated R code:

```{r eval=FALSE,tidy=FALSE}
## This file produces a plot relating the explanatory variable to the outcome.
thedata <- read.csv("Data/thedata-15-03-2011.csv") ## Read the data
pdf("fig1.pdf") ## begin writing to the pdf file
please-plot(outcome by explanatory using thedata. red lines please.)
please-add-a-line(using model1)
## Note to self: a quadratic term does not add to the substance
## model2 <- please-fit(outcome by explanatory+explanatory^2 using thedata
## summary(abs(fitted(model1)-fitted(model2)))
dev.off() ## stop writing to the pdf file
```

Now, in the future if you wonder how 'that plot on page 10' was created, you
will know: (1) 'that plot' is from a file called `fig1.pdf`and (2)
`fig1.pdf`was created in `makefig1.R`. Any changes to the file in the future (provided R still
exists), will just require some quick edits of commands already
written.^[We use data from @norris2015data throughout this paper.]  Even if in the future R ceases to exist, you (or someone else) will at
least be able to read the plain text commands and use them to write code in a
new favorite statistical computing language: R scripts are written in [plain
text](http://en.wikipedia.org/wiki/Plain_text), and plain text is a format that
will be around as long as computer programmers write computer programs.

Moreover, coding saves time. Often, a lot of time, especially if projects grow
in the number of files, partners, or complexity of analysis. Manually importing
one data file into R, may be effortless. Importing 100 files is another issue,
and the time costs of each manual action add up quickly (and the probability
for error increases too).

Also, realize that *file names send messages* to coauthors and to your future self. This means
that if you name your files with evocative and descriptive names, your
collaborators are less likely to call you at midnight asking for help and you
will remove some regret from your future self and protect your friendships and
working relationships. For example, if you are studying inequality and protest,
you might try naming a file something like `inequality-and-protest-figures.R`
instead of `temp9.R` or `supercalifragilisticexpialidocious.R`.  By the way,
the extension `.R` tells us and the operating system that the file contains R
commands. This part of the filename enables us to quickly search our antique
hard drives for files containing R scripts.^[We think that some method of
tagging files by the purpose of the file will continue help analysts find and organize their files for long after the idea of a computer mouse ceases to make sense.]

Coding helps us to avoid making mistakes. For example, in our example above we may
be interested in how many people protest. We may use a data file containing all
protests for several years. People often create a tabular display of this data
by copying the results manually to the working paper document. In copying we
can make mistakes. So a better approach is to automatically create the table (in the format you like, with
horizontal lines, 3 decimals, a title, etc) and output this in whatever file
type you need (.tex. .pdf, .rtf, etc). Now when we obtain new data a new table can be created
quickly, so we make less mistakes and save time!

For example, the following table was created entirely with code using R and the `xtable` package @dahl2016aa.\footnote{@beck2010reg inspired this particular
presentation of a linear model.}

```{r loadlibraries, echo=FALSE}
makebecktable<-function(lmobj,vars,thecaption,thelabel,filename){
	require(xtable)
	##  The next lines extract relevant and useful output from the
	##  regression object, following Neal Beck's TPM article 2010.
	themat<-cbind(summary(lmobj)$coef[,c("Estimate","Std. Error")],confint(lmobj))
	n<-length(lmobj$fitted.values)
	r2<-signif(summary(lmobj)$r.squared,2)
	est.sigma<-signif(summary(lmobj)$sigma,2)
	colnames(themat)[3:4]<-c("~{}","~{}")
	row.names(themat)<-vars
	thextab<-xtable(themat,align=c("l",rep(".",4)),digits=1,
			caption=thecaption,
			label=thelabel)
	print(thextab,
	      file=filename,
	      include.colnames=FALSE,
	      hline.after=NULL,
	      table.placement="H",
	      comment=FALSE,
	      sanitize.text.function=function(x){x},
	      add.to.row=list(
			      pos=list(0,3),
			      command=c(
					"\\toprule \n &
					\\multicolumn{1}{c}{Coef} &
					\\multicolumn{1}{c}{Std. Err.} &
					\\multicolumn{2}{c}{95\\% CI} \\\\ \\cmidrule(r){2-5} ",
				paste("& \\multicolumn{4}{c}{n: ",n,",",
				      "   resid.sd: ",signif(est.sigma),
				      ", R$^2$: ",r2,"} \\\\ \\bottomrule",sep="")
				)
			      ))

@@

```{r maketable1, echo=TRUE, tidy=FALSE}
lm1<-lm(protest05~gini04+meanpr,data=good.df)  #  Run the regression
makebecktable(lmobj=lm1, # make the table file
	      vars=c("Intercept","Income Inequality (lower=more equal)",
		     "Mean Political Rights (lower=more rights)"),
	      thecaption="People living in countries with unequal
	      income distributions report more protest activity to World Values
	      Survey interviewers than people living in countries with relatively
	      more equal income distributions, adjusting for average political
	      rights as measured by Freedom House 1980--2010.",
	      thelabel="tab:protest",
	      filename="protesttable.tex")
```

```
\input{protesttable.tex} ## read in the table file
```

@@

**Step 1** Code everything that can be coded. If we know the provenance of results, future or current collaborators make less mistakes and can quickly and easily reproduce (and thus change and improve) upon the work.

@@

Data analysis involves a long series of decisions. Each decision requires a
justification.  Some decisions will be too small and technical for inclusion in
the published article itself. Still, these need to be documented in the code
itself [@nagler1995coding]. Paragraphs and citations in the publication will
justify the most important decisions but the code itself documents the smaller
but still consequential decisions. So, one must code to communicate with
yourself and others. There are two main ways to avoid forgetting the reasons
you did something with data: comment your code and tightly link your code with
your writing -- making your code literate.

@@

How might one avoid these problems?  [Literate
programming](http://en.wikipedia.org/wiki/Literate_programming) is the practice
of weaving code into a document --- paragraphs, equations, and diagrams can
explain the code, and the code can produce numbers, figures, and tables (and
diagrams and even equations and paragraphs). Literate programming is not merely
fancy commenting but is about enabling the practice of programming itself to
facilitate easy reproduction and communication.

For example, in [@sec:data-analys-comp], we suggested that we knew where 'that
plot on page 10' comes from by making sure we had a `fig1.pdf` file produced
from a clearly commented plain text file called something like `makefig1.R`.
An even easier solution would be to directly include a chunk of code to produce
the figure inside of the paper itself. This paper, for example, was written in
plain text using markdown markup with R code chunks to make things like Figure
\ref{fig:giniprot}. This combination of markdown and R is called [R
Markdown](http://rmarkdown.rstudio.org). Here, is an example of embedding a
plot inside of a document. In a document sent to a journal one would tend to hide the code chunks and thus one would only see the plot.

```{r fig1plotcode,eval=FALSE,echo=TRUE}
##  Make a scatterplot of Protest by Inequality
par(bty="n",xpd=TRUE,pty="s",tcl=-.25)
with(good.df,plot(gini04,protest05,
                  xlab='Gini Coefficient 2004 (UNDP)',
                  ylab='Proportion Reporting Protest Activities\n(World Values Survey 2005)',
                  cex=.8))
##  Label a few interesting points
with(good.df[c("Brazil","United Kingdom","United States","Sweden","Chile"),],
     text(gini04,protest05,labels=Nation,srt=0,cex=.6,pos=3,offset=.1))
```

```{r giniprot,message=FALSE,echo=FALSE,fig=TRUE,include=TRUE,width=4,height=4,fig.cap="Average number of protest activities by income inequality across countries in 2004--2005.",out.width=".5\\textwidth",fig.env="figure",fig.lp="fig:",fig.pos="H",fig.align="center"}
##  Make a scatterplot of Protest by Inequality
par(bty="n",xpd=TRUE,pty="s",tcl=-.25)
with(good.df,plot(gini04,protest05,
                  xlab='Gini Coefficient 2004 (UNDP)',
                  ylab='Proportion Reporting Protest Activities\n(World Values Survey 2005)',
                  cex=.8))
##  Label a few interesting points
with(good.df[c("Brazil","United Kingdom","United States","Sweden","Chile"),],
     text(gini04,protest05,labels=Nation,srt=0,cex=.6,pos=3,offset=.1))
```

Markdown (and LaTeX and HTML) all have ways to cross-reference within a
document. For example, by using `\label{fig:giniprot}` in \LaTeX or
`{#fig:giniprot}` in the pandoc version of markdown, we do not need to keep
track of the figure number, nor do extra work when we reorganize the document
in response to reviewer suggestions. Nor do we need a separate `makefig1.R`file
or `fig1.pdf` file.  Tables and other numerical results are also possible to
generate within the source code of a scholarly paper. For example, if we had
ommitted the filename in `makebecktable` above, we would have seen a LaTeX
table be generated within this document itself. For Stata users, there now is
[Markdoc](https://github.com/haghish/MarkDoc) which is very similar to R
markdown.

The R project has a task view devoted to [reproducible
research](http://cran.r-project.org/web/views/ReproducibleResearch.html)
listing many of the different approaches to literate programming for R.  If
your workflow does not involve R, you can still implement some of the
principles here. Imagine creating a style in Microsoft Word called 'code 'which
hides your code when you print your document, but which allows you to at least
run each code chunk piece by piece (or perhaps there are ways to extract all
text of style code from a Microsoft Word document). Or imagine just using some
other kind of indication linking paragraphs to specific places in code files.
There are many ways that creative people can program in a literate way.

Literate programming need not go against the principle of modular data analysis
[@nagler1995coding]. Jake routinely uses several different files that fulfill
different functions, some of them create \LaTeX code that he can `\input` into
his `main.tex` file, others setup the data, run simulations, or allow him to
record his journeys down blind alleys. Of course, when we have flying cars
running on autopilot, perhaps something other than combining R and Markdown or
LaTeX will make our lives even easier. Then we'll change.

**Step 2** We analyze data in order to explain something about the world to
other scholars and policy makers. If we focus on explaining how we got our
computers to do data analysis, we will do a better job with the data analysis
itself: we will learn as we focus on teaching others about why we did what we
did, and we will avoid errors and save time as we ensure that others (including
our future selves) can retrace our steps. A document than can be 'run' to
reproduce all of the analyses also instills confidence in readers and can more
effectively spur discussion and learning and cumulation of research.

@@

```{bash ,eval=FALSE, tidy=FALSE}
00_archive/
01_paper/
  20160622_inequality.tex
02_data/
  00_archive/
  01_rawdata/
  02_cleandata/
  20160622_analysis.dta
  20160622_analysis.rda
03_analysis/
  00_archive/
  01_temp/
  02_output/
    01_tables/
    02_figures/
  20160622_analysis.do
  20160622_analysis.R
  20160622_prep_data.do
  20160622_prep_data.R
 ```

@@

```{bash, tidy=FALSE, eval=FALSE}
manuscript\
figures\
analysis\
  README.md
data\
build\
libraries\
README.md
Makefile
```

@@

# Minimize error by testing.

@@

```{r results='asis'}
test.times.2.fn()
```

@@

```{r}
smalldat <- subset(good.df,subset=protest05<90)
stopifnot(all(smalldat$protest05<90))
```

@@

# Create a reproducible workflow

>>>>

---
title: How to improve your relationship with your future self
author:
- name: Jake Bowers
  affiliation: <jwbowers@illinois.edu>,  University of Illinois at Urbana-Champaign, Departments of Political Science and Statistics; Fellow, White House Social and Behavioral Sciences Team.
- name: Maarten Voors
  affiliation: <maarten.voors@wur.nl>, Wageningen University, Development Economics Group.  \emph{Acknowledgments:} Many thanks to the EGAP Learning Days 2016 participants in Santiago Chile, to the BITSS team, the Department of Economics and Ted Miguel at UC Berkeley where Maarten was a visiting researcher during spring 2016. Maarten gratefully acknowledges financial support from N.W.O. grant 451-14-001. A previous version of this paper benefited from comments and discussions with Mark Fredrickson, Brian Gaines, Kieran Healy, Kevin Quinn, Cara Wong, Mika LaVaque-Manty and Ben Hansen. The source code for this document may be freely downloaded and modified from <https://github.com/jwbowers/workflow>. This paper extends a previous version by Jake @bowers2011tpm. While most of the text remains the same we have expanded and updated the essay to reflect current developments in technology and thinking about transparency and reproducibility of research.
bibliography: references.bib
csl: american-political-science-review.csl
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
    pdf_document:
      keep_tex: true
      number_sections: true
      fig_width: 7
      fig_height: 7
      fig_caption: true
      template: bowersarticletemplate.latex
      pandoc_args: [
      "-M", "secPrefix=\\S",
      "--filter", "/usr/local/bin/pandoc-crossref"
      ]
      md_extensions: +autolink_bare_uris+ascii_identifiers+tex_math_single_backslash
...

<!-- An example comment -->
<!-- Separate the thanks sections from affiliation and integrate? Wasn't sure how to do this here-->

@@

## To make the html file do
## render("exploration4.Rmd", output_format=html_document(fig_retina=FALSE))
## To make the pdf file do
## render("exploration4.Rmd", output_format=pdf_document())
## to debug, try render("thefile.Rmd", clean=FALSE) so that you can see the intermediate files

require(knitr)
opts_chunk$set(
  tidy = TRUE,     # display code as typed
  size = "footnotesize",    # slightly smaller font for code
  echo = TRUE,
  results = 'markup',
  strip.white = TRUE,
  fig.path = 'figs/fig',
  cache = FALSE,
  highlight = TRUE,
  tidy.opts=list(width.cutoff = 60),
  out.width = '.9\\textwidth',
  message = FALSE,
  comment = NA)
```

```{r initialize, echo = FALSE}
##First, just setup the R environment for today:
if(!file.exists('figs')) dir.create('figs')

options(SweaveHooks = list(fig = function(){
			   par(mar = c(3.5, 3, 1.1, 0),
			       pty = "s",
			       mgp = c(1.5,0.5,0),
			       oma = c(0,0,0,0))},
			 echo = function(){options(continue = " ") ##Don't show "+" prompts,
			 options(prompt = " ")
			 }),
	digits = 4,
	scipen = 8,
	width = 80
	)

#options(error = function(){options(prompt = "> ", continue = "+ ");NULL})
```

@@

Memory is tricky. Learning requires effort. When we do not practice and repeat something that we want to remember, most people forget quickly, are overconfident in their abilities to recall future information [@koriat2005illusions], and may even recall events that never happened.^[See the following site for a nice overview of what we know about memory --- including the fact that learning requires practice: <http://www.spring.org.uk/2012/10/how-memory-works-10-things-most-people-get-wrong.php>. On false memory see Wikipedia and linked studies <https://en.wikipedia.org/wiki/False_memory>.] Moreover, most of us live busy lives. We type text knowing that the laundry needs doing, hearing children play or cry, ignoring the news or the latest journal review. Our lives and minds are full and we need to efficiently move from task to task. If we cannot count on memory, then how can we do science?

How long does it take from planning a study to publication and then to the first reproduction of it? Is three years too short? Is ten years too long? We suspect that few of our colleagues in the social and behavioral sciences conceive of, field, write, and publish a data driven study faster than about three years. We also suspect that, if some other scholar decides to re-examine the analyses of a published study, it will occur after publication. Moreover, this new scholarly activity of learning from one another's data and analyses can occur at anytime, many years past the initial publication of the article.^[The process of reproducing past findings can occur when one researcher wants to build on the work of another. It can also occur within the context of classes --- some professors assign reproduction tasks to students to aid learning about data analysis and statistics. In addition to those models, reproduction of research has recently been organized to enhance the quality of public policy in the field of economic development by the 3IE Replication Program (@brown2014quality) and to assess the quality of scientific research at within social psychology (@open2015estimating) and within experimental economics (@camerer2016evaluating). In another study, 29 research teams recently collaborated on a project focusing on applied statistics to see if the same answers would emerge from re-analyses of the same data set (@silberzahn2015crowdsourced). They didn't.]

@@

```{r read.data, echo = FALSE, results = 'hide', cache = TRUE}
## This data frame will be used throughout the analyses.
## Putting it up top, so we can use it in the comments section.
library(readstata13)
dem.nations.df <- read.dta13("https://www.dropbox.com/s/huhqag5rudwtbno/Democracy%20Cross-National%20Data%20V4.1%2009092015.dta?dl=1&pv=1&raw=1", convert.factors = FALSE)

row.names(dem.nations.df) <- dem.nations.df$Nation ##make the row names more helpful
good.df <- na.omit(dem.nations.df[, c("Gini2004","WVS5_protest","Nation","meanpr")])
good.df$gini04 <- good.df$Gini2004/100
good.df$protest05 <- good.df$WVS5_protest
```

We organize the paper around a series of homilies that lead to certain concrete actions:

 - Data analysis is computer programming.
 - No data analyst is an island for long.
 - Meaningful code requires data.
 - Version control prevents clobbering, reconciles history, and helps organize work.
 - Testing minimizes error.
 - Work can be reproducible.
 - Research ought to be credible communication.

@@

All your results (numbers, comparisons, tables, plots, figures) should be produced from code, not from a series of mouse clicks or copying and pasting.^[To our future selves: both of us are using a computer control device called a track-pad, and haven't used an older device called a [mouse](https://en.wikipedia.org/wiki/Computer_mouse) in some years. Our current track-pads do not click. We are not sure why we talked about mouse clicks just now.] Imagine you wanted to re-create a figure and include a new variable, you should be able to do so with just a few edits to the code rather than knowledge of how you used a pointing device in your graphical user interface all those years ago.

Lets look at an example. Using an open-source statistics programming language called R [@R:2016], you
might specify that a file, called `fig1.pdf` is produced by the following set
of commands in a file called `makefig1.R`. Lets look at some annotated R code:

```{r eval = FALSE, tidy = FALSE, highlight = FALSE}
# This file produces a plot relating the explanatory variable to the outcome.
## Read the data
thedata <- read.csv("Data/thedata-15-03-2011.csv")
## begin writing to the pdf file
please-open-pdf("fig1.pdf")
please-plot(outcome by explanatory using thedata. red lines please.)
please-add-a-line(using model1)
## Note to self: a quadratic term does not add to the substance
## model2 <- please-fit(outcome by explanatory+explanatory^2 using thedata
## summary(abs(fitted(model1)-fitted(model2)))
## stop writing to the pdf file
please-close-pdf()
```

Now, in the future if you wonder how 'that plot on page 10' was created, you
will know: (1) 'that plot' is from a file called `fig1.pdf`and (2)
`fig1.pdf`was created in `makefig1.R`. Any changes to the file in the future
will just require some quick edits of commands already written  (provided R
still exists).^[We use data from @norris2015data throughout this paper.]  Even
if in the future R ceases to exist, you (or someone else) will at least be able
to read the plain text commands and use them to write code in a new favorite
statistical computing language: R scripts are written in [plain
text](http://en.wikipedia.org/wiki/Plain_text), and plain text is a format that
will be around as long as computer programmers write computer programs.

Moreover, coding saves time. Often, typing commands into a file saves a lot of
time, especially if projects grow in the number of files, collaborators, or
complexity of analysis. Manually importing one data file into R may be
effortless. Importing 100 files is another issue, and the time costs of each
manual action add up quickly (and the probability for mistakes increases too).

Also, realize that **file names send messages** to coauthors and to your future self. This means
that if you name your files with evocative and descriptive names, your
collaborators are less likely to call you at midnight asking for help and you
will remove some regret from your future self and protect your friendships and
working relationships. For example, if you are studying inequality and protest,
you might try naming a file something like `inequality-and-protest-figures.R`
instead of `temp9.R` or `supercalifragilisticexpialidocious.R`.  By the way,
the extension `.R` tells us and the operating system that the file contains R
commands. This part of the filename enables us to quickly search our antique
hard drives for files containing R scripts.^[We think that some method of
tagging files by the purpose of the file will continue help analysts find and organize their files for long after the idea of a computer mouse ceases to make sense.]

Coding helps us to avoid making mistakes. For example, in our example above we may
be interested in how many people protest. We may use a data file containing all
protests for several years. People often create a tabular display of this data
by copying the results manually to the working paper document. In copying we
can make mistakes. So a better approach is to automatically create the table (in the format you like, with
horizontal lines, 3 decimals, a title, etc) and save this table in whatever file
type you need (.tex. .pdf, .rtf, etc). Now when we obtain new data, a new table can be created
quickly, so we make less mistakes and save time!

For example, the following table was created entirely with code using R and the `xtable` package @dahl2016aa.\footnote{@beck2010reg inspired this particular presentation of a linear model.} We then read this table into the current file using the LaTeX command `\input{protesttable.tex}` which produces a nice looking table in pdf format.

```{r loadlibraries, echo = FALSE}
makebecktable <- function(lmobj, vars, thecaption, thelabel, filename){
        require(xtable)
        ##  The next lines extract relevant and useful output from the
        ##  regression object, following Neal Beck's TPM article 2010.
        themat <- cbind(summary(lmobj)$coef[, c("Estimate","Std. Error")], confint(lmobj))
        n <- length(lmobj$fitted.values)
        r2 <- signif(summary(lmobj)$r.squared,2)
        est.sigma <- signif(summary(lmobj)$sigma,2)
        colnames(themat)[3:4] <- c("~{}","~{}")
        row.names(themat) <- vars
        thextab <- xtable(themat, align = c("l", rep(".",4)), digits = 1,
                          caption = thecaption,
                          label = thelabel)
        print(thextab,
              file = filename,
              include.colnames = FALSE,
              hline.after = NULL,
              table.placement = "H",
              comment = FALSE,
              sanitize.text.function = function(x){x},
              add.to.row = list(
                                pos = list(0,3),
                                command = c(
                                            "\\toprule \n &
                                            \\multicolumn{1}{c}{Coef} &
                                            \\multicolumn{1}{c}{Std. Err.} &
                                            \\multicolumn{2}{c}{95\\% CI} \\\\ \\cmidrule(r){2-5} ",
                                    paste("& \\multicolumn{4}{c}{n: ", n,",",
                                          "   resid.sd: ", signif(est.sigma),
                                          ", R$^2$: ", r2,"} \\\\ \\bottomrule", sep = "")
                                    )
                                ))

@@

```{r maketable1, echo = TRUE, tidy = TRUE, tidy.opts=list(width=50)}
#  Run the regression
lm1 <- lm(protest05~gini04+meanpr, data = good.df)
# make the table file
makebecktable(lmobj = lm1,
              vars = c("Intercept","Income Inequality (lower = more equal)",
                       "Mean Political Rights (lower = more rights)"),
              thecaption = "People living in countries with unequal
              income distributions report more protest activity to World Values
              Survey interviewers than people living in countries with relatively
              more equal income distributions, adjusting for average political
              rights as measured by Freedom House 1980--2010.",
              thelabel = "tab:protest",
              filename = "protesttable.tex")
```

@@

Notice that we might have made one file called `makefig1.R` and another called `protestincomedisttab.R` which in turn produce a pdf file (`fig1.pdf`) and a LaTeX format table (`protesttable.tex`). This idea of modularity [@nagler1995coding] in code enables us to quickly find errors, and, if we want to make changes to improve our work, only one file may be changed at a time. Further, it enhances collaboration --- Jake can work to make the figure and Maarten can work to make the table without worry about conflicting files.

### Step 1
Code everything that can be coded. If we know the provenance of results, future or current collaborators make less mistakes and can quickly and easily reproduce (and thus change and improve) upon the work.

@@

Data analysis involves a long series of decisions. Each decision requires a
justification.  Some decisions will be too small and technical for inclusion in
the published article itself. Still, these need to be documented in the code
itself [@nagler1995coding]. Paragraphs and citations in the publication will
justify the most important decisions but the code itself documents the smaller
but still consequential decisions. So, one must code to communicate with
yourself and others. There are two main ways to avoid forgetting the reasons
you did something with data: comment your code and tightly link your code with
your writing---making your code literate.

@@

How might one avoid these problems?  [Literate
programming](http://en.wikipedia.org/wiki/Literate_programming) is the practice
of weaving code into a document such that paragraphs, equations, and diagrams can
explain the code, and the code can produce numbers, figures, and tables (and
diagrams and even equations and paragraphs). Literate programming is not merely
fancy commenting but is about enabling the practice of programming itself to
facilitate easy reproduction and communication.

For example, in [@sec:data-analys-comp], we suggested that we knew where 'that
plot on page 10' comes from by making sure we had a `fig1.pdf` file produced
from a clearly commented plain text file called something like `makefig1.R`.
An even *easier* solution would be to directly include a chunk of code to
produce the figure inside of the paper itself. This paper, for example, was
written in plain text using markdown markup with R code chunks to make things
like Figure \ref{fig:giniprot}.^[This combination of markdown and R is called
[R Markdown](http://rmarkdown.rstudio.org).] Here is an example of embedding a
plot inside of a document along with the source of this paragraph. In a
document sent to a journal one would tend to hide the code chunks and thus one
would only see the plot.

```
For example, in [@sec:data-analys-comp], we suggested that we knew where
'that plot on page 10' comes from by making sure we had a `fig1.pdf` file
produced from a clearly commented plain text file called something like
`makefig1.R`.  An even *easier* solution would be to directly include a chunk
of code to produce the figure inside of the paper itself. This paper, for
example, was written in plain text using markdown markup with R code chunks to
make things like Figure \ref{fig:giniprot}.^[This combination of markdown and R
is called [R Markdown](http://rmarkdown.rstudio.org).] Here is an example of
embedding a plot inside of a document along with the source of this
paragraph. In a document sent to a journal one would tend to hide the code
chunks and thus one would only see the plot.
```

```{r fig1plotcode, eval = FALSE, echo = TRUE}
##  Make a scatterplot of Protest by Inequality
par(bty = "n", xpd = TRUE, pty = "s", tcl = -.25)
with(good.df, plot(gini04, protest05,
                  xlab = 'Gini Coefficient 2004 (UNDP)',
                  ylab = 'Proportion Reporting Protest Activities\n(World Values Survey 2005)',
                  cex = .8))
##  Label a few interesting points
with(good.df[c("Brazil","United Kingdom","United States","Sweden","Chile"),],
     text(gini04, protest05, labels = Nation, srt = 0, cex = .6, pos = 3, offset = .1))
```

```{r giniprot, message = FALSE, echo = FALSE, fig = TRUE, include = TRUE, width = 4, height = 4, fig.cap = "Average number of protest activities by income inequality across countries in 2004--2005.", out.width = ".5\\textwidth", fig.env = "figure", fig.lp = "fig:", fig.pos = "H", fig.align = "center"}
##  Make a scatterplot of Protest by Inequality
par(bty = "n", xpd = TRUE, pty = "s", tcl = -.25)
with(good.df, plot(gini04, protest05,
                  xlab = 'Gini Coefficient 2004 (UNDP)',
                  ylab = 'Proportion Reporting Protest Activities\n(World Values Survey 2005)',
                  cex = .8))
##  Label a few interesting points
with(good.df[c("Brazil","United Kingdom","United States","Sweden","Chile"),],
     text(gini04, protest05, labels = Nation, srt = 0, cex = .6, pos = 3, offset = .1))
```

Markdown (and LaTeX and HTML) all have ways to cross-reference within a
document. For example, by using `\label{fig:giniprot}` in LaTeX or
`{#fig:giniprot}` in the [pandoc](http://pandoc.org/) version of markdown built
into [RStudio](https://www.rstudio.com/), we do not need to keep track of the
figure number, nor do extra work when we reorganize the document in response to
reviewer suggestions. Nor do we need a separate `makefig1.R`file or `fig1.pdf`
file.  Tables and other numerical results are also possible to generate within
the source code of a scholarly paper. For example, if we had omitted the
filename in `makebecktable` above, the LaTeX formatted table would have appeared
within this document itself. For Stata users, there now is
[Markdoc](https://github.com/haghish/MarkDoc) which is very similar to R
markdown.

The R project has a task view devoted to [reproducible
research](http://cran.r-project.org/web/views/ReproducibleResearch.html)
listing many of the different approaches to literate programming for R.  If
your workflow does not involve R, you can still implement some of the
principles here. Imagine creating a style in Microsoft Word called 'code' which
hides your code when you print your document, but which allows you to at least
run each code chunk piece by piece (or perhaps there are ways to extract all
text of style 'code' from a Microsoft Word document for use in some other
program). Or one could just use some other kind of indication linking
paragraphs to specific places in code files.  There are many ways that creative
people can program in a literate way.

Literate programming need not go against the principle of modular data analysis
[@nagler1995coding]. Jake routinely uses several different files that fulfill
different functions, some of them create LaTeX code that he can `\input` into
his `main.tex` file, others setup the data, run simulations, or allow him to
record his journeys down blind alleys. Of course, when we have flying cars
running on autopilot, perhaps something other than combining R and Markdown or
LaTeX will make our lives even easier. Then we'll change.

### Step 2
We analyze data in order to explain something about the world to
other scholars and policy makers. If we focus on explaining how we got our
computers to do data analysis, we will do a better job with the data analysis
itself: we will learn as we focus on teaching others about why we did what we
did, and we will avoid errors and save time as we ensure that others (including
our future selves) can retrace our steps. A paper than can be "run" to
reproduce all of its analyses also instills confidence in readers and can more
effectively spur discussion and learning and cumulation of research.

@@

```{bash , eval = FALSE, tidy = FALSE}
00_archive/
01_paper/
  20160622_inequality.tex
02_data/
  00_archive/
  01_rawdata/
  02_cleandata/
  20160622_analysis.dta
  20160622_analysis.rda
03_analysis/
  00_archive/
  01_temp/
  02_output/
    01_tables/
    02_figures/
  20160622_analysis.do
  20160622_analysis.R
  20160622_prep_data.do
  20160622_prep_data.R
 ```

@@

```{bash, tidy = FALSE, eval = FALSE}
manuscript\
figures\
analysis\
  README.md
data\
build\
libraries\
README.md
Makefile
```

@@

# Testing minimizes error.

@@

```{r results = 'asis'}
test.times.2.fn()
```

@@

```{r}
smalldat <- subset(good.df, subset = protest05<90)
stopifnot(all(smalldat$protest05<90))
```

@@

# Work can be reproducible.
